Transmembrane helices are also an overlooked source of major histocompatibility complex class II epitopes and evolutionary more conserved than expected by chance

### 2020-06-23 bbbq_article.tex
Improved Methods quite a bit

Also for MHC-II, there are multiple computational tools developed 
to predict epitopes,
such as using a trained neural network (\cite{nielsen2003reliable})
or a Gibbs sampling approach (\cite{nielsen2004improved}).
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCIIpan (\cite{nielsen2008quantitative,karosiene2013netmhciipan})
gave rise to the most accurate predictions.
Later tools are \cite{zhang2013predicting},
\cite{trolle2014nettepi}, \cite{zhang2015accurate}
and the very recent MHCnuggets\cite{shao2020high}.
In this study, we'll be using MHCnuggets, as it is the only FOSS tool available.

@@

There are multiple computational tools developed to detect which
parts of membrane proteins are TMH
In 2001, multiple tools to do so have been compared \cite{moller2001evaluation},
of which TMHMM \cite{krogh2001predicting} turned out to be the best.
Many other tools followed, such as 
Phobius (\cite{kall2004combined}),
ConPred II (\cite{arai2004conpred}),
MEMSAT3 (\cite{jones2007improving})
and MetaTM (\cite{klammer2009metatm}), which signals the importance
of TMH predictions. 
Unlike TMHMM, which is still up and running, 
these later tools have already become obsolete.
Contemporary tools that are good for use, such as
the closed-source MemBrain \cite{feng2020topology} 
and the FOSS tool PureseqTM \cite{wang2019efficient}.
In this study, we will use TMHMM  for pragmatic reasons
\richel{I would love to use PureseqTM instead, but only if there is time}.

@@

If presentation of TMHs on MHC-I would bring an evolutionary advantage 
in the recognition of pathogens by the immune system, 
it would follow that this is equally important for MHC-II, 
especially as the help of CD4+ T cells is needed for a long lasting CD8+ T cell 
response \cite{novy2007cd4}. 
The mechanism to detect foreign TMHs by MHC-II would be unknown, 
similar to the discovery that MHC-I presents TMHs.

@@

The human immune system and human pathogen are in an evolutionary
arms race: our immune systems is selected for the detection
of pathogens, whereas pathogens are selected to avoid detection.
From a pathogen's point of view, however, this struggle 
is of only minor importance:
in seasonal influenza, for example, the selection pressure
exerted by the immune system was only limited \cite{han2019individual}

@@

To determine the percentages of epitopes overlapping
with TMHs, we use two reference proteomes.
Our viral AA sequence is represented by
the proteome of the first sequenced COVID-19 strain (\cite{wu2020new},
GenBank ID of MN908947.3, \url{https://www.ncbi.nlm.nih.gov/nuccore/MN908947})
For bacteria, we use the reference genome of 
Mycobacterium tuberculosis (\url{https://www.ebi.ac.uk/reference_proteomes}, UP000001584, 
83332 MYCTU).

\paragraph{Data sets for evolutionary selection}

To detect an evolutionary signal for selection on binding
TMHs peptides, we need a phylogenetic tree with 
many time-dated proteomes.
For humans, we use the 25000 proteomes used in influenza \richel{citation here}.
For viruses, we use the many time-dated COVID-19 proteomes.
We can assume that only after the moment that COVID-19 spilled over 
to humans, it has mostly been the human immune system selecting against
its detection, although this is only part of the selection for 
having a high transmission rate.
For bacteria, use the proteomes of Mycobacterium 
\richel{Really? Pick better set, if possible!}.

To determine the percentages of MHC-I epitopes overlap
with TMHs, we used the same 
analysis as \cite{bianchi2017},
except for using a viral and bacterial proteome (instead 
of a human proteome, as used in the original study).
Bianchi and colleagues obtained a distribution of 
percentages of MHC-I epitopes overlap with TMHs in Homo sapiens
for the different HLA haplotypes, with an average of 5.3\%.
We obtained a similar distribution of percentages of MHC-I epitopes that 
overlap with TMHs for the different HLA haplotypes, but then applied to
our viral and bacterial proteome.

We compare the distributions of humans and each pathogen
using a two-sample Kolmogorov-Smirnov (KS) test
for a significance level $\alpha = 0.05$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{MHC-2}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To determine the percentages of MHC-II epitopes overlap, 
we do the same analysis as MHC-I, except for
MHC-II haplotypes. We picked the MHC-II alleles
that occur with a frequency of at least 20\% in
the human population (\cite{greenbaum2011functional},
see \ref{table:mhc2_haplotypes} for coverage),
resulting in \richel{approx 8} HLAs.
The threshold of 20\% is arbitrarily, yet chosen to result in 5 and 10 HLA variants.  
As \cite{bianchi2017}, we define the 5\% of peptides 
with the lowest IC50 values as binders. We then counted the percentage
of binders that have one or more amino acids that are part of a TMH.

@@

\paragraph{Gene prediction}

Proteogenomics is the field dedicated to, among others,
predict the proteins that arise from a DNA 
sequence \cite{nesvizhskii2014proteogenomics}.
For an ab initio gene prediction of a viral genome, 
we searched the literature for the most recent software tool,
which is Vgas (\cite{zhang2019vgas}). In this article, 
Vgas is compared to tools, such as GeneMarkS (\cite{besemer2001genemarks}),
Prodigal (\cite{hyatt2012gene}), GLIMMER (\cite{delcher1999improved}),
RAST \cite{aziz2008rast} and prokka (\cite{seemann2014prokka}).
We picked the FOSS tool that scored best in the (closed source) Vgas
benchmark (\cite{zhang2019vgas}), which is Prodigal.

@@

Because we use 13 MHC-I and 7 MHC-II haplotypes
for a significance level of 5\%, by chance we expect (5\% of $13 + 7$)
one times to find a (MHC-I or MHC-II) haplotype 
that is suggested to have selection working upon it.
Applying a binomial distribution ($p = 0.05$, $n = 13+7$, $\alpha = 0.05$),
we will claim we there is some form of selection when we find 3
or more examples of having a significant result.
However, because there are 2 hypotheses tested per haplotype,
we will claim there is directional selection, 
when, for a same hypothesis, we find 2 examples of having a significant result.

% > pbinom(q = 2, size = 20, prob = 0.05, lower.tail = FALSE)
% [1] 0.07548367
%
% pbinom(q = 3, size = 20, prob = 0.05, lower.tail = FALSE)
% [1] 0.01590153

>>>>

Also for MHC-II, there are multiple computational tools developed 
to predict epitopes,
such as using a trained neural network (\cite{nielsen2003reliable})
or a Gibbs sampling approach (\cite{nielsen2004improved}).
According to \cite{lundegaard2011prediction}, in 2011,
from a set of multiple tools, 
NetMHCIIpan (\cite{nielsen2008quantitative,karosiene2013netmhciipan})
gave rise to the most accurate predictions.
Later tools are \cite{zhang2013predicting},
\cite{trolle2014nettepi}, \cite{zhang2015accurate}
and the very recent MHCnuggets (\cite{shao2020high}).
In this study, we'll be using MHCnuggets, as it is the only FOSS tool available.

@@

There are multiple computational tools developed to detect which
parts of membrane proteins are TMH
In 2001, multiple tools to do so have been compared \cite{moller2001evaluation},
of which TMHMM \cite{krogh2001predicting} turned out to be the best.
Many other tools followed, such as 
Phobius (\cite{kall2004combined}),
ConPred II (\cite{arai2004conpred}),
MEMSAT3 (\cite{jones2007improving})
and MetaTM (\cite{klammer2009metatm}), which signals the importance
of TMH predictions. 
Contemporary tools are the closed-source MemBrain (\cite{feng2020topology}) 
and the FOSS tool PureseqTM (\cite{wang2019efficient}).
In this study, we will use PureseqTM.

@@

If presentation of TMHs on MHC-I would bring an evolutionary advantage 
in the recognition of pathogens by the immune system, 
it would follow that this is equally important for MHC-II, 
especially as the help of CD4+ T cells is needed for a long lasting CD8+ T cell 
response (\cite{novy2007cd4}). 
The mechanism to detect foreign TMHs by MHC-II would be unknown, 
similar to the discovery that MHC-I presents TMHs.

@@

The human immune system and human pathogen are in an evolutionary
arms race: our immune systems is selected for the detection
of pathogens, whereas pathogens are selected to avoid detection.
From a pathogen's point of view, however, this struggle 
is of only minor importance:
in seasonal influenza, for example, the selection pressure
exerted by the immune system was only limited (\cite{han2019individual}).

@@

To determine the percentages of epitopes overlapping
with TMHs, we use three reference proteomes, which are
a human, viral and bacterial one.
The human reference proteome is \richel{here}, 
which is the same as \cite{bianchi2017}.
The viral reference proteome is represented by
the proteome of the first sequenced COVID-19 strain (\cite{wu2020new},
GenBank ID of MN908947.3, \url{https://www.ncbi.nlm.nih.gov/nuccore/MN908947})
For bacteria, the reference genome of 
Mycobacterium tuberculosis (\url{https://www.ebi.ac.uk/reference_proteomes}, UP000001584, 
83332 MYCTU) is used.

\paragraph{Data set for evolutionary selection}

To detect an evolutionary signal for selection on binding
TMHs peptides, we need a phylogenetic tree with 
many time-dated proteomes.
To do so, we use a viral pathogen, as these are detected by both
MHC-I and MHC-II.
To detect individual mutations, we need a high resolution dataset
that also includes the evolutionary history of the virus.
COVID-19 satisfies both requirements:
as of 2020-06-22, it has 49790 proteomes, as provided by GISAID (\cite{shu2017gisaid}),
as well as a provided evolutionary history by Nextstrain (\cite{hadfield2018nextstrain}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Measuring TMH epitopes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To determine the percentages of MHC-I and MHC-II epitopes overlap
with TMHs, we used mostly the same analysis as described in \cite{bianchi2017}.
To summarize: a proteome is converted into all its possible 9-mers. For each
of these peptides, it is measured if it has been part of a TMH (at least
1 residue), as well as if the peptide is a binder to an MHC-I haplotype.
Because the percentage of TMH residues is known, a binomial test can be
used to determine if TMH-derived peptides are as likely to be a binder
as expected by chance (the null hypothesis), or that TMHs are indeed an
overrepresented source of epitopes (as found by \cite{bianchi2017}).

This study differs in some aspects, described below in more detail,
which is the definition of what a binder is,
the addition of using MHC-II haplotypes, the software used,
and the significance level to determine if TMH-derived peptides are as 
likely to be a binder as expected by chance.
These deviations are either a refinement of the previous method or
a pragmatic choice made due to the extension of the original experiment.

The definition of a binder differs from \cite{bianchi2017}:
in this study a peptide is labeled an MHC-I binder if, within a haplotype, 
any of its 9-mer peptides have an IC50 in the lowest 2\% of randomly simulated 
peptides (see Supmat for values \richel{Put these values in supmat}), were the original study defined
a binder as having an IC50 in the lowest 2\% of the peptides within a protein.
% See https://github.com/richelbilderbeek/bianchi_et_al_2017/blob/72e6755a31d400158368509fd80a41e984677ab1/predict-binders.R#L17

The 13 MHC-I haplotypes used in this study are the same as 
the ones in \cite{bianchi2017}.
The MHC-II haplotypes used additionally are selected 
to occur with a phenotypic frequency of at least 14\% in
the human population (\cite{greenbaum2011functional},
see \ref{table:mhc2_haplotypes} for coverage),
resulting in 13 haplotypes.
This frequency threshold is arbitrarily chosen to result in 
the same number of haplotypes for each MHC class
\richel{Add this to bbbq, there it is now 21}.
When using an MHC-II haplotype, instead of using 9-mers, 13-mers are
used, as these are the canonical MHC-II epitope size \richel{Reference here}.

Instead of using the MHC-I-only epitope prediction 
tool \verb;EpitopePrediction;, MHCnuggets is used, as it can predict
both MHC-I and MHC-II epitopes.
Additionally, instead of only using a human proteome, in this study
also a viral and bacterial proteome are analyzed.

Where the original study used a significance level of 1\% 
for the binomial test to determine if TMH-derived peptides 
are as likely to be a binder as expected by chance, this study
uses 5\%, which is a common significance level for an exploratory study.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comparing amounts of TMH epitopes}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To determine if MHC-I has the same percentage of epitopes overlapping
with viral/bacterial TMHs as with human TMHs \richel{$\mathcal{H}_{1,VB}$}, 
we compare the distributions of the measured percentages 
of TMH-overlapping peptides for all haplotypes,
and compare the distributions between humans and each pathogen
using a two-sample Kolmogorov-Smirnov (KS) test
for a significance level $\alpha = 0.05$, 
which is a common significance level for an exploratory study.

@@

% \paragraph{Gene prediction}
% 
% Proteogenomics is the field dedicated to, among others,
% predict the proteins that arise from a DNA 
% sequence \cite{nesvizhskii2014proteogenomics}.
% For an ab initio gene prediction of a viral genome, 
% we searched the literature for the most recent software tool,
% which is Vgas (\cite{zhang2019vgas}). In this article, 
% Vgas is compared to tools, such as GeneMarkS (\cite{besemer2001genemarks}),
% Prodigal (\cite{hyatt2012gene}), GLIMMER (\cite{delcher1999improved}),
% RAST \cite{aziz2008rast} and prokka (\cite{seemann2014prokka}).
% We picked the FOSS tool that scored best in the (closed source) Vgas
% benchmark (\cite{zhang2019vgas}), which is Prodigal.

@@

Because we use 13 haplotypes per MHC class,
for a significance level of 5\%, 
there is a 14% chance to find one haplotype 
that hints to have selection working upon it.
% > pbinom(q = 1, size = 13, prob = 0.05, lower.tail = FALSE)
% [1] 0.1354239

The chance to find 2 haplotypes that hint to have selection working upon it
is 2\%, which is below our significance level.
% > pbinom(q = 2, size = 13, prob = 0.05, lower.tail = FALSE)
% [1] 0.02450784

Therefore, we will claim there is directional selection, 
when, for a proteome, we find 2 haplotypes 
that hint to have selection working upon it
